---
title: "Getting started with cvpa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cvpa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

cvpa gives you weighted party support and policy-issue attitudes by
demographic group, over time, from Canadian public-opinion surveys going
back to 1945. As of version 0.2.0, that covers several hundred policy
questions across seven issue domains -- economy and taxation, energy and
environment, health/education/social programs, immigration/diversity/
federalism, foreign policy and identity, social issues/rights/justice, and
democracy and institutions -- in addition to vote choice and intention.
From cvpa's point of view, vote is just one more question: it's decoded and
analyzed the same way as any policy question.

## Two ways to use cvpa

There are two separate paths through the package, and it's worth being
clear about which one you're on:

* **The live data path** (`update_cvpa_data()`, `question_catalog()`,
  `question_data()`, `wtd_response()`, `response_gap_analysis()`) reads
  Parquet files hosted alongside the [cp3 web
  app](https://www.quantoid.ca/files/app-js-remote/), covering every
  question in the catalog. This is almost certainly what you want for new
  analysis.
* **The legacy vote path** (`vote_data`, `wtd_vote()`, `gap_analysis()`)
  is the original vote-choice/intention dataset and functions the package
  shipped with before 0.2.0, kept as-is for reproducing prior work (in
  particular Lucas, Romualdi, Armstrong, and Goodyear-Grant's *Canadian
  Journal of Political Science* piece on the Canadian Vote Intention
  Dataset). `wtd_vote()`/`gap_analysis()` are, under the hood, thin
  wrappers around the same general engine that powers the live data path,
  so their behavior hasn't changed -- see the last section of this
  vignette.

This vignette focuses on the live data path, since that's what's new.

```{r setup}
library(cvpa)
```

## Setting up the local cache

cvpa doesn't bundle policy-attitude data in the package itself -- there's
too much of it, and it's updated periodically as new survey waves are
added. Instead, `update_cvpa_data()` downloads what you ask for into a
local cache (see `cvpa_cache_dir()`) the first time you need it, and every
other function reads only from that cache. Nothing in the package other
than `update_cvpa_data()` ever touches the network, so you're always in
control of when a download happens.

The question catalog and the demographic/party code-labels table are small
and are always refreshed by `update_cvpa_data()`. Respondent-level data for
a specific question is only fetched when you name it (or ask for
`questions = "all"`, which is on the order of 90MB across ~630 questions,
as of this writing):

```{r update-data}
update_cvpa_data(questions = "bri_econ_013")
```

*(The code chunks below assume this succeeded. If you're reading this
vignette somewhere without network access, the code is shown for reference
but its output won't appear.)*

```{r have-data, include = FALSE}
have_data <- requireNamespace("nanoparquet", quietly = TRUE) &&
  tryCatch({
    update_cvpa_data(questions = "bri_econ_013", quiet = TRUE)
    TRUE
  }, error = function(e) FALSE)
knitr::opts_chunk$set(eval = have_data)
```

## Finding a question

`question_catalog()` lists every available question, one row per question,
deduplicated across the several hundred survey waves each question has
accumulated over the decades:

```{r catalog}
question_catalog(domain = "econ")
```

`domain` and `type` (`"b"` for binary questions, `"lsm"` for 3-point
less/same/more-style questions, `"nsa"` for 3-point never/sometimes/always-
style questions, or `"party"` for vote choice/intention) both filter the
catalog if you want to narrow it down. With no arguments, `question_catalog()`
returns everything, vote included:

```{r catalog-all}
nrow(question_catalog())
table(question_catalog()$domain)
```

Once you've picked a question, `question_response_labels()` shows you what
its response categories actually are and how they're coded:

```{r response-labels}
question_response_labels("bri_econ_013")
```

## Fetching and inspecting a question's data

`question_data()` reads the (now cached) respondent-level data for one
question and, by default, decodes both the response and every demographic
column into labeled factors:

```{r question-data}
d <- question_data("bri_econ_013")
head(d)
```

Every question's data has the same shape: `response`, `weight`, `year`,
plus the demographic and vote columns listed in `cvpa_grouping_vars` --
see `demographic_vars` for what each demographic variable measures and
what categories it takes:

```{r grouping-vars}
cvpa_grouping_vars
demographic_vars
```

Notice that `vote_intention` and `vote_choice` are in that list alongside
the demographics -- every question's data carries how each respondent
voted, so you can group any policy question by vote just as easily as by
gender or region.

## Weighted response shares over time

`wtd_response()` computes each response category's weighted share of the
total, for every year (and, optionally, grouping-variable combination):

```{r wtd-response}
wtd_response(d, grouping_vars = "gender")
```

A quick look at how the "More" category (support for reducing the
rich-poor gap) has moved over time, by gender:

```{r plot-wtd-response, fig.width = 6, fig.height = 4}
more_by_year <- subset(wtd_response(d, grouping_vars = "gender"), response == "More")
# reshape() matches explicitly on year rather than assuming the two
# genders' rows line up positionally -- safer if you adapt this to a
# question/grouping variable that doesn't have every level in every year.
wide <- reshape(as.data.frame(more_by_year[c("year", "gender", "response_share")]),
                 idvar = "year", timevar = "gender", direction = "wide")
matplot(
  wide$year, wide[, -1],
  type = "l", lty = 1, lwd = 2, col = c("steelblue", "firebrick"),
  xlab = "Year", ylab = "Share answering \"More\"",
  main = "Support for reducing the rich-poor gap, by gender"
)
legend("bottomright", legend = c("Man", "Woman"), col = c("steelblue", "firebrick"), lty = 1, lwd = 2, bty = "n")
```

## Testing whether a gap is significant

`response_gap_analysis()` generalizes `gap_analysis()` to any question: a
difference-of-proportions test between two levels of a grouping variable,
for each response category, for each year:

```{r response-gap-analysis}
response_gap_analysis(d, grouping_var = "gender")
```

The `flag` column tells you whether every cell going into that year's test
had an expected count above 5 (`"p-val reliable"`) or not
(`"p-val unreliable"`) -- see `make_flag()`.

## Vote is just another question

Because vote choice and intention live in the same catalog and the same
`resp/question=<code>/data_0.parquet` layout as every policy question,
everything above works identically for `"vote_choice"`/`"vote_intention"`:

```{r vote-as-question}
update_cvpa_data(questions = "vote_choice")
vc <- question_data("vote_choice", years = 2000:2021)
wtd_response(vc, grouping_vars = "gender")
```

## The legacy vote path

If you're working with the original `vote_data` dataset -- for instance to
reproduce results from the CJPS piece it was built for -- `wtd_vote()` and
`gap_analysis()` are unchanged:

```{r legacy}
data(vote_data)
wtd_vote(vote_data, "intention", years = 2000:2022, grouping_vars = "gender")
gap_analysis(vote_data, "intention", years = 2000:2022, grouping_var = "gender")
```

Note that `vote_data` predates the cp3 Parquet pipeline and has a single
`community_size` column rather than the live data path's `com_100`/`com_500`
split -- see `demographic_vars` for the live path's version of each
demographic variable.
